# Creating publication-grade figures {#sec-ggplot}

## General philosophy

In science, we want clear and informative plots. Each figure should make it obvious what data you are plotting, what the axes, colors, shapes, and size differences represent, and the overall message the figure is conveying. When writing a scientific paper or report, remember that your future readers are busy people. They often do not have the time to delve into the subtleties of overly refined verbal arguments. Instead, they will most often look for the figures to learn what your work is about. You will want to create figures which makes this possible for them to do.

Here we will learn how to create accessible, publication-quality scientific graphs in a simple way. We do this using the R package `ggplot2` which is a standard part of the `tidyverse`. The `ggplot2` package follows a very special philosophy for creating figures that was originally proposed by Leland @Wilkinson2006. The essence of this view is that, just like the grammar of sentences, graphs have fixed "grammatical" components whose specification defines the plot. The grand idea is that the data ought not be changed in order to display it in different formats. For instance, the same data should be possible to represent either as a box plot, or as a histogram, without changing their format.

This last claim needs to be qualified somewhat. It is more accurate to say that one should not need to change the data *as long as they are in tidy format*. As a reminder, "tidy data" means that every variable is in its own column, and every observation is in its own row (@sec-tidy). In case the data are not tidy, one should first wrangle them into such form, for example by using `pivot_longer`. While this step is not always required (especially for simpler graphs), it can be very useful to tidy the data before analyzing and plotting them when working with larger, more complex datasets.


## Basic `ggplot2` usage

To see how `ggplot2` works, let us load `tidyverse`, and then use the built-in `iris` dataset to create some figures. As a reminder, here is what the data look like:

```{r}
#| message: false
library(tidyverse)
as_tibble(iris)
```

Let us, as a first step, create a plot where sepal length (x-axis) is plotted against petal length (y-axis), with the points referring to different species shown in different colors: 

```{r}
ggplot(
  data = iris,
  mapping = aes(x = Sepal.Length, y = Petal.Length, color = Species)
) +
  geom_point()
```

The `ggplot` function takes two inputs: the **data** (in this case, the `iris` table) and the **aesthetic mappings** (`mapping`). On top of this, we add on the **geometry** of how to display the data---in this case, with `geom_point()`.

The aesthetic mappings are always defined via the `aes` helper function. What are these "aesthetic mappings"? The important thing to remember is that the aesthetic mappings are all those aspects of the figure that are governed by the data. For instance, if you wanted to set the color of all points to blue, this would not be an aesthetic mapping, because it applies regardless of what the data are (in case you want to do this, you would have to specify `geom_point(color = "blue")` in the last line).

The **geometry** of your plot governs the overall visual arrangement of your data (points, lines, histograms, etc). There are many different `geom_`s; we will learn about some here, but when in doubt, Google and a [`ggplot2` cheat sheet](https://nyu-cdsc.github.io/learningr/assets/data-visualization-2.1.pdf) are your best friends.

The above code is often written by piping the data into the `ggplot` function. Then, since the `mapping` argument is always given via the `aes` function, one usually omits naming this argument:

```{r}
iris |>
  ggplot(aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point()
```

is the most common form one writes the function call to `ggplot`.

To look at a different kind of geometry, let us create a histogram of the petal lengths. This is done using `geom_histogram`:

```{r}
#| message: false
iris |>
  ggplot(aes(x = Petal.Length)) +
  geom_histogram()
```

Before looking at the result, notice that the different "grammatical" components of the plot (aesthetics, geometry) are *added* to the plot, using the `+` symbol for addition. A common source of error is to accidentally keep using the pipe operator `|>` even within a plot. The rule of thumb is that after invoking `ggplot`, one must use `+` to compose the various graph elements, but outside of that, the usual `|>` is used for function composition. If one uses `|>` instead of `+` within a plot, R will give back an error message instead of graphics:

```{r}
#| error: true
iris |>
  ggplot(aes(x = Petal.Length)) |>
  geom_histogram()
```

Just be sure to remember this so you can correct the mistake, should you accidentally run up against it.

Let us now look at the plot above. We see two clusters of data. Why is that? One might suspect that this is because of a species-level difference. To check if that is the case, let us color the histogram by species:

```{r}
#| message: false
iris |>
  ggplot(aes(x = Petal.Length, color = Species)) +
  geom_histogram()
```

(A color legend was created automatically on the right.) This changes the color of the outline of the histograms, but not the color of their fill. To do so, we need to change the `fill` color as well, which is a separate aesthetic property:

```{r}
#| message: false
iris |>
  ggplot(aes(x = Petal.Length, color = Species, fill = Species)) +
  geom_histogram()
```

This is fine, but now the histogram bars of different species are stacked on top of one another. This means that, for example at around a petal length of 5 cm where individuals of both *Iris versicolor* and *I. virginica* are found, the green bar on top of the blue one does not begin from the bottom of the y-axis at 0, but from wherever the blue bar ends and the green one begins (in our case, from around 10 upwards). So the number of *I. versicolor* individuals in the 5 cm bin is not 12, but only 2.

It may be easier to interpret the histogram if all bars start at the bottom, even if this means that the bars will now overlap to an extent. To achieve this overlap, one simply has to add the argument `position = "identity"` to `geom_histogram`:

```{r}
#| message: false
iris |>
  ggplot(aes(x = Petal.Length, color = Species, fill = Species)) +
  geom_histogram(position = "identity")
```

(There are other options as well, such as `position = "dodge"` -- feel free to experiment. In general, you can learn what options each of the `geom_`s have by taking a look at their help pages, or by asking Google.) The figure above is almost perfect; the one remaining problem is that the bars belonging to different species cover each other completely. So let us change the fill's *transparency*, which is called `alpha` in ggplot. The `alpha` property can take on any value between 0 (fully transparent object, to the point of invisibility) to 1 (fully solid object with no transparency, like above; this is the default). Setting it to 0.2 (say) will finally fully reveal the distribution of each species:

```{r}
#| message: false
iris |>
  ggplot(aes(x = Petal.Length, color = Species, fill = Species)) +
  geom_histogram(position = "identity", alpha = 0.2)
```

There are plenty of other `geom_`s as well, such as `geom_boxplot`, `geom_violin`, etc. Let us try `geom_boxplot` for instance. Let us create one box plot of the distribution of petal lengths for each species. Putting the species along the x-axis and petal length along the y-axis:

```{r}
iris |>
  ggplot(aes(x = Species, y = Petal.Length)) +
  geom_boxplot()
```

Although this is sufficient, one should feel free to make the plots prettier. For instance, one could use colors, like before. One can also use different themes. There are pre-defined themes such as `theme_classic()`, `theme_bw()`, and so on. For example, using `theme_bw()` gets rid of the default gray background and replaces it with a white one. The version of the graph below applies this theme, and also changes the color and fill to be based on species identity:

```{r}
iris |>
  ggplot(aes(x = Species, y = Petal.Length,
             color = Species, fill = Species)) +
  geom_boxplot(alpha = 0.2) +
  theme_bw()
```


## Summaries and confidence intervals {#sec-CI}

Providing appropriate information on experimental errors is a hallmark of any credible scientific graph. Choose a type of error based on the conclusion that you want the reader to draw. For example, one can show the largest and smallest values in the dataset apart from the average.

Here we will display the average petal length of each species together with the largest and smallest values. To do so, we first obtain the mean, maximum, and minimum of the petal lengths for each species. We then plot them. So we first `group_by` the species and `summarize` to obtain the required statistics (means and standard deviations), and hand these data over for plotting. One can include all these steps in a logical workflow:

```{r}
#| message: false
iris |>
  group_by(Species) |> # Perform summary calculations for each species
  summarize(mP = mean(Petal.Length), # Mean petal length
            minP = min(Petal.Length), # Minimum petal length
            maxP = max(Petal.Length)) |> # Maximum petal length
  ungroup() |> # Ungroup data
  # Start plotting:
  ggplot(aes(x = Species, y = mP, ymin = minP, ymax = maxP)) +
  geom_point() + # This takes the y aesthetic, for plotting the mean
  geom_errorbar(width = 0.2) # Takes the ymin and ymax aesthetics
```

Note that the y-axis label is not very descriptive. This is because it inherited the name of the corresponding data column, `mP`. There are multiple ways to change it; the simplest is to add `labs(y = "Petal length")` to the plot. Another way of doing so is discussed in @sec-moreggplot.

Another way of characterizing uncertainty is to turn away from the spread of values in the actual raw data and instead focus on the uncertainty of an estimated quantity such as the mean. The rest of this subsection shows how to calculate confidence intervals for estimated means, and is intended for those who are already familiar with basic statistical concepts.

In case we want to calculate the 95% confidence intervals of the mean values, we first obtain some necessary summary statistics: the number of observations (sample size) in each group; the standard error of the mean (standard deviation divided by the square root of the sample size); and finally, the confidence interval itself (read off from the *t*-distribution, with one fewer degrees of freedom than the sample size). We can then include these confidence intervals on top of the mean values:

```{r}
#| message: false
iris |>
  group_by(Species) |>
  summarize(mP = mean(Petal.Length), # Mean petal length per species
            sP = sd(Petal.Length), # Standard deviation per species
            N = n(), # Sample size (number of observations) per species
            SEM = sP / sqrt(N), # Standard error of the mean
            CI = SEM * qt(1 - 0.025, N - 1)) |> # Confidence interval,
                 # read off at 1 - 0.025 because +/- 2.5% adds up to 5%
  ggplot(aes(x = Species, y = mP, ymin = mP - CI, ymax = mP + CI)) +
  geom_point() +
  geom_errorbar(width = 0.2) +
  ylab("Petal length")
```

In the literature, it is common to encounter bar-and-whisker plots to represent the same information as above. These replace the points showing the means with bars that start from zero. One should be aware of how to read and make such graphs. This is almost trivially simple: all one needs to do is replace `geom_point` with `geom_col`.

```{r}
#| message: false
iris |>
  group_by(Species) |>
  summarize(mP = mean(Petal.Length),
            sP = sd(Petal.Length),
            N = n(),
            SEM = sP / sqrt(N),
            CI = SEM * qt(1 - 0.025, N - 1)) |>
  ggplot(aes(x = Species, y = mP, ymin = mP - CI, ymax = mP + CI)) +
  # Below we change geom_point to geom_col (for "column"):
  geom_col(alpha = 0.4, color = "steelblue", fill = "steelblue") +
  geom_errorbar(width = 0.2, color = "steelblue") +
  ylab("Petal length")
```

:::{.callout-warning}
While means and their confidence intervals are often displayed with bar-and-whisker plots, there is an important reasons why this is not a good idea. By starting the bars from zero, the plot implies that zero is a natural point of comparison for all the data. Unfortunately, this can visually distort the information we wish to convey. Consider the following graph:

```{r}
tibble(Species = c("species 1", "species 2"),
       Average = c(148, 152),
       CI = c(0.8, 0.9)) |>
  ggplot(aes(x = Species, y = Average,
             ymin = Average - CI, ymax = Average + CI)) +
  geom_col(alpha = 0.4, color = "steelblue", fill = "steelblue") +
  geom_errorbar(width = 0.2, color = "steelblue") +
  ylab("Trait value")
```

It is impossible to see whether there are any relevant differences between the two species. The following is exactly the same, but with the mean values shown with points instead of bars:

```{r}
tibble(Species = c("species 1", "species 2"),
       Average = c(148, 152),
       CI = c(0.8, 0.9)) |>
  ggplot(aes(x = Species, y = Average,
             ymin = Average - CI, ymax = Average + CI)) +
  geom_point(color = "steelblue") +
  geom_errorbar(width = 0.2, color = "steelblue") +
  ylab("Trait value")
```

It is now obvious that the two observations are distinct. Due to this problem, my recommendation is to avoid bar-and-whisker plots and simply use point-and-whisker plots instead.
:::


## Exercises {#sec-ggplot-exercises}

@Fauchaldetal2017 tracked the population size of various herds of caribou in North America over time, and correlated population cycling with the amount of vegetation and sea-ice cover. The part of their data that we will use consists of two files: [`pop_size.tsv`](https://raw.githubusercontent.com/dysordys/data-with-R/main/data/pop_size.zip) (data on herd population sizes), and [`sea_ice.tsv`](https://raw.githubusercontent.com/dysordys/data-with-R/main/data/sea_ice.zip) (on sea levels of sea ice cover per year and month).

1.  The file `sea_ice.tsv` is in human-readable, wide format. Note however that the rule "each set of observations is stored in its own row" is violated. We would like to organize the data in a tidy tibble with four columns: `Herd`, `Year`, `Month`, and `Cover`. To this end, apply the function `pivot_longer` to columns 3-14 in the tibble, gathering the names of the months in the new column `Month` and the values in the new column `Cover`.

2.  Use `pop_size.tsv` to make a plot of herd sizes through time. Let the x-axis be Year, the y-axis be population size. Show different herds in different colors. For the geometry, use points.

3.  The previous plot is actually not that easy to see and interpret. To make it better, add a `line` geometry as well, which will connect the points with lines.

4.  Make a histogram out of all population sizes in the data.

5.  Make the same histogram, but break it down by herd, using a different color and fill for each herd.

6.  Instead of a histogram, make a density plot with the same data and display (look up `geom_density` if needed).

7.  Make box plots of the population size of each herd. Along the x-axis, each herd should be separately displayed; the y-axis should be population size. The box plots should summarize population sizes across all years.

8.  Let us go back to `sea_ice.tsv`. Make the following plot. Along the x-axis, have `Year`. Along the y-axis, `Month`. Then, for each month-year pair, color the given part of the plot darker for lower ice cover and lighter for more. (Hint: look up `geom_tile` if needed.) Finally, make sure to do all this only for the herd with the label `WAH` (filter the data before plotting).

The following plotting exercises use the Galápagos land snail data (@sec-snail).

9.  Create a density plot (`geom_density`) with shell size along the x-axis and the corresponding density of individuals along the y-axis. How much overlap is there between the sizes of the different species?

10. Repeat the above exercise, but use the standardized size instead of the raw measurements. As a reminder, this means subtracting out the minimum from each entry, and dividing the results by the difference of the maximum and minimum entries (@sec-mutate). Make sure to have clean, descriptive axis labels on the plot. What is the difference between this figure and the one obtained in the previous exercise using the non-standardized size values?

11. Now do exercises 9-10 for shell shape instead of size.

12. Create a scatter plot with `size` along the x-axis and `shape` along the y-axis. Each individual should be a point on the graph.

13. Repeat the same, but use the standardized `size` and `shape` measurements along the axes. Compare the figures. In what do they differ?

14. Create the same scatter plot with standardized `size` along the x-axis and `shape` along the y-axis, but with the points colored based on `habitat`. That is, individuals found in `humid` environments should all have one color, and those found in `arid` regions should have another. Do you see any patterns, in terms of whether certain shell sizes or shapes are more associated with a given type of habitat?

15. Modify the previous plot by coloring the points based on species instead of habitat. How much trait overlap is there between individuals belonging to different species? How do you interpret this result, especially in light of what you previously saw in exercises 9-11?
